---
title: "step 2: smoothing navigation data"
author: "Your Name"
date: "Today's Date"
format:
  html:
    toc: true
    number-sections: true
editor: 
  markdown: 
    wrap: sentence
---

## Setup

Before beginning, we will load necessary libraries.

```{r}


library(lubridate)
library(tsibble)
library(zoo) # for interpolations
library(plotly) # for interactive plots 
library(tidyverse)
library(magrittr) 
library(htmlwidgets) # for creating interactive html plots
library(mgcv)



getwd() -> wd
# delete the "code" part of the pathway so that the rest of the path are not relative to the code directory
wd <- str_remove(wd, "/code") 

# set the working directory to the directory above "code"
setwd(wd)

# list the csv and txt tables and print their names.
list.files(paste0( wd,"/annotations"), pattern = ".csv$" ) 
# look for files that end with .csv or .txt 
list.files(paste0( wd,"/nav"), pattern = ".csv$|.txt$" ) %>% 

# exclude the "arranged" pattern - it will be used in files geerated by this code
str_subset("smoothed", negate = T)  %>% 
  paste(., collapse = "\n - ") %>% 
  paste0("Avialable CSV tables: \n - ",.,"\n") %>% cat()

```

# Load data

open the table with time and xy positions,  

```{r}
# set your navigation file name, 
navigation_file <- "navigation_ROV16.txt"
print(paste0("loading navigation file: ", navigation_file))
read_csv(paste0(wd,"/nav/",navigation_file) ) -> navigation

# order columns for consistency - discard the ones you wont use
navigation %>% select(datetime, videotime,lat, lon,  depth) -> navigation

# show what navigation you are loading 
navigation %>% head() %>%  print()
 
```

# set the time column

depending on the formate of your time column and The mood of Excel, you might need to change the time column to a datetime object if it is not done automatically

```{r}

 # fill the gaps in the shp file
# time management
navigation %>% 
# make the time column - if you have a data and a time column
  mutate(datetime = datetime) -> positions
# if you have a dmy_hms character string


# check that datetime is a datetime object
# if class of datetime is not "POSIXct" or "POSIXt" then convert it
if(class(positions$datetime)[1] %in% c("POSIXct", "POSIXt")) {
  "datetime is valid" %>% print()
}else if(!class(positions$datetime)[1] %in% c("POSIXct", "POSIXt"))  {
  print("setting datetime to POSIXt")
  positions %>%  mutate( datetime = dmy_hms(datetime)) -> positions
} else { # send a warning if this fails. 
  # check what format your time column is in 
positions %>% 
  select(datetime) %>% glimpse
    stop("cannot read the date time column make sure it can be read correctly in lubridate::dmy_hms() function")
}

# make a time column of elapsed seconds based on column datetime, in case for if videotime (logged via ROV system) is not starting at 0
positions$elapsed_seconds <- as.numeric(difftime(positions$datetime, positions$datetime[1],units = "secs"))
```


# Smooth

*mgcv::gam* function specifically fits a Generalized additive model (GAM) using P-splines and smoothing parameter estimation method by REML

```{r}

# Smooth the track  

# Extract the variables to be smoothed from the navigation
t <- positions$elapsed_seconds # input variable is time in seconds
x <- positions$lon
y <- positions$lat
z <- positions$depth


### Fit a GAM with p-spline and smoothing parameter estimation method based on REML to each dimension. 

gam_east_gcv <- gam(x ~ s(t, bs = "ps", k=175),  method="REML")
gam_north_gcv <- gam(y ~ s(t, bs = "ps", k=175),  method="REML")
gam_depth_gcv <- gam(z ~ s(t, bs = "ps", k=175),   method="REML") 

### predict 
smooth_x <- predict(gam_east_gcv   ) %>% as.vector()

smooth_y <- predict(gam_north_gcv   ) %>% as.vector()

smooth_z <- predict(gam_depth_gcv   )%>% as.vector()



```


Assessment of basis dimension k. EDF and k index should be interpret carefully. For further reading, see mgcv help.

```{r}


# GAM smoothing model assessment
gam.check(gam_east_gcv) %>% print()
gam.check(gam_north_gcv) %>% print()
gam.check(gam_depth_gcv) %>% print()



```



# set frequency to 1s and predict postion at missing time stamps

using the GAM models to predict the position with a new set of time stamps at a given frequency (by default, 1s)

The zoo::na.approx function is part of the zoo package in R, which is designed for handling time series data and other ordered observations. The na.approx function is specifically used to approximate or interpolate missing values (NAs) in a numeric vector or time series. It assumes a linear relationship between the known data points surrounding the missing values to estimate them.

```{r}

# add the new coordinates to the table 
tibble( xsmoothed = smooth_x ,
        ysmoothed = smooth_y ,
        depthsmoothed = smooth_z )  %>%
  bind_cols(positions,.) %>%
  select(datetime, videotime,lon,lat, depth, elapsed_seconds,xsmoothed, ysmoothed,depthsmoothed )  -> smoothed_navigation
 
# time interval check - if there are missing time stamps, create them and predict the coordinates for those time stamps as well.
if(all(diff(smoothed_navigation$elapsed_seconds) == 1)) {
    print("No missing time stamps, no need to fill gaps")
  smoothed_navigation_filled <- smoothed_navigation
# create the missing time stamps so that you have a reading at each second   
} else if(any(diff(smoothed_navigation$elapsed_seconds) == 1)) {
  print("missing 1s intervals - filling up the gaps" )
  smoothed_navigation %>% 
    as_tsibble(index = datetime) %>%
  fill_gaps() %>% # automatically creates the missing timestamps at 1s interval
  as_tibble() -> smoothed_navigation_filled

# if the time interval is larger than 1s, create the missing time stamps at 1s interval and predict the coordinates for those time stamps as well.
} else if(all(diff(smoothed_navigation$elapsed_seconds) > 1) ){ 
  print("setting time increment to 1s and filling up the gaps" )
  smoothed_navigation  %>% 
  add_row(. , datetime  = .$datetime[1] + 1  , .before = 2)  %>% 
    as_tsibble(index = datetime) %>%
  fill_gaps() %>% # automatically creates the missing timestamps at 1s interval
  as_tibble() -> smoothed_navigation_filled
  
} else {
stop("Time interval is not consistent, please check your data and make sure it is in 1s intervals or has a consistent time interval")
}


# make a new time vector for seconds
t2  <- as.numeric(difftime(smoothed_navigation_filled$datetime, smoothed_navigation_filled$datetime[1],units = "secs"))



# add t2 as videotime in seconds in 1 second intervals
smoothed_navigation_filled %>% 
 mutate(LON2 = predict(gam_east_gcv, newdata = data.frame(t = t2) ) %>% as.vector() ,
        LAT2 = predict(gam_north_gcv, newdata = data.frame(t = t2) ) %>% as.vector(), 
        DEPTH2 = predict(gam_depth_gcv, newdata = data.frame(t = t2) ) %>% as.vector(),
        videotime_interpolated = t2 %>% as.vector()
 ) -> interpolated_smoothed_navigation



# check that there are no NA left in LON2 or LAT2 or DEPTH2
if(any(is.na(interpolated_smoothed_navigation$LON2)) | any(is.na(interpolated_smoothed_navigation$LAT2)) | any(is.na(interpolated_smoothed_navigation$DEPTH2))) {
  print("There are NA values in LON2, LAT2 or DEPTH2")
} else {
  print("No NA values in LON2, LAT2 or DEPTH2. That's great")
}
 



```

# plot the interpolated smoothed track
```{r}


plot3d <- plot_ly() %>%
  # make smaller point size
  add_trace( data = interpolated_smoothed_navigation,
             x = ~lon, y = ~lat, z = ~(-depth),
             mode = "markers", type = "scatter3d", name = "un-smoothed",
             marker = list(size = 3,color = "blue4"), 
             # display x y z and datetime when hovering over each point
             text = ~datetime%>% as.character(),
             hovertemplate = paste('<b>Lon</b>: %{x}',
                                   '<br><b>Lat</b>: %{y}',
                                   '<br><b>Depth</b>: %{z}',
                                   '<br><b>Time</b>: %{text}',sep = "" )
  ) %>%
  # use the smoothed XYZ to plot the track
  add_trace( data = interpolated_smoothed_navigation,
             x = ~lon, y = ~lat, z = ~(-depth),
             mode = "line", type = "scatter3d", name = "Track - un-smoothed ",
             line = list(size = 4, color = "cyan") , 
             # display x y z and datetime when hovering over each point
             text = ~datetime%>% as.character(),
             hovertemplate = paste('<b>Lon</b>: %{x}',
                                   '<br><b>Lat</b>: %{y}',
                                   '<br><b>Depth</b>: %{z}',
                                   '<br><b>Time</b>: %{text}',sep = "" )
  ) %>%
  
  # add the GAM smoothed
    add_trace(x = interpolated_smoothed_navigation$LON2,
              y = interpolated_smoothed_navigation$LAT2,
              z = - interpolated_smoothed_navigation$DEPTH2,
              mode = "markers", type = "scatter3d", name = "GAM smoothed with p-spline ",
              marker = list(size = 3, color = "tomato"), 
              # display x y z and datetime when hovering over each point
              text = ~datetime%>% as.character(),
              hovertemplate = paste('<b>Lon</b>: %{x}',
                                    '<br><b>Lat</b>: %{y}',
                                    '<br><b>Depth</b>: %{z}',
                                    '<br><b>Time</b>: %{text}',sep = "" )
    )  %>% 
    # plot hte smoothed track as well 
  add_trace(x = interpolated_smoothed_navigation$LON2,
            y = interpolated_smoothed_navigation$LAT2,
            z = - interpolated_smoothed_navigation$DEPTH2,
            mode = "lines", type = "scatter3d", name = "Track - GAM smoothed with p-spline",
            line = list(size = 4, color = "orange"), 
            # display x y z and datetime when hovering over each point
            text = ~datetime%>% as.character(),
            hovertemplate = paste('<b>Lon</b>: %{x}',
                                  '<br><b>Lat</b>: %{y}',
                                  '<br><b>Depth</b>: %{z}',
                                  '<br><b>Time</b>: %{text}',sep = "" )
  ) %>%
 
  # plot parameters 
  layout(title = "Difference of un-smoothed and smoothed & interpolated ROV transect",
         scene = list(xaxis = list(title = "Longitude"),
                      yaxis = list(title = "Latitude"),
                      zaxis = list(title = "Depth")))  

# view the plot locally 
plot3d %>% print()  

# export it as a standalone object that you can share without the rest of the code
#  
saveWidget(plot3d, paste0(wd, "/Output/figures/",
                          navigation_file %>% str_remove(pattern = ".csv$|.txt$"),
                          "3D Navigation_steep3 ", ".html") )

print(paste0("3D plot of smoothed and interpolated navigation data saved as: ", paste0(wd, "/Output/figures/",
                          navigation_file %>% str_remove(pattern = ".csv$|.txt$"),
                          "3D Navigation_steep3", ".html") ) )


```


# Export

Attach the smoothed XY and the interpolated depth to the navigation

```{r}


write_csv( interpolated_smoothed_navigation , paste0(wd,"/nav/smoothed_",navigation_file))   


```
