---
title: "step 4: Biigle video 3D distance travelled"
author: "Your Name"
date: "Today's Date"
format:
  html:
    toc: true
    number-sections: true
editor: 
  markdown: 
    wrap: sentence
---

## Setup

Before beginning, we will load necessary libraries.

Also input the parameters (resolution) of the images and the laser scale in meters. 

```{r}
library(sf)
library(rgl)
library(magrittr)
library(tidyverse)

getwd() -> wd
# delete the "code" part of the pathway so that the rest of the path are not relative to the code directory
wd <- str_remove(wd, "/code") 

# set the working directory to the directory above "code"
#setwd(wd)

 

```

# Load data

open the table with time and xy positions, here maybe need to change/delete working direction???
(linked to my local drive) setwd("D:/PHD/ROV/ROV raw video density estimates/navigation_smoothing")

```{r}
# put path to your processed navigation file with smoothed XYZ and image width
paste0(wd,"/Output/laserCal_biigleannotation.csv" ) %>% read_csv() -> smoothed_navigation_laser_data

```

# 

```{r}

# if you know it, put the corresponding filename here
 # - Check if needed !!!!
smoothed_navigation_laser_data %>% mutate(filename = "generic_transect" ) %>%
  # put the filename in first place
  relocate(filename,.before = names(smoothed_navigation_laser_data)[1])

# use the video time interpolated to merge metdata and navigation data  
# this should be where the start time is   
smoothed_navigation_laser_data %>% filter( videotime_interpolated == 0 ) %>% pull(datetime)
 

#  checks the start_time is 0 seconds in videos
if(smoothed_navigation_laser_data %>% slice(1) %>% pull(elapsed_seconds) == 0 ) {
  print("start time is 0 seconds in video, good to go!")
  video_starttime <- smoothed_navigation_laser_data %>% slice(1) %>% pull(datetime)
} else {
  print("start time is not 0 seconds in video, check your data and make sure the start time is correct for calculating distance travelled")
  smoothed_navigation_laser_data %>% slice(1) %>% pull(datetime) -> v1
  smoothed_navigation_laser_data %>% slice(1) %>% pull(elapsed_seconds) -> v2
  video_starttime <- v1 - v2 # calculate the start time by subtracting the elapsed seconds from the datetime of the first row
  
  }

 
time_stamps <- smoothed_navigation_laser_data  %>%  
  mutate(sec_in_video = seconds(datetime  -  video_starttime   )  %>% seconds_to_period()  ) %>% 
  mutate(time_in_video =sprintf('%02d:%02d:%02d', sec_in_video$hour,sec_in_video$minute, sec_in_video$second) )

# change format of time column in navigation table
time_stamps %<>% 
  mutate( sec_in_video =  sec_in_video %>% seconds() )

# does this look legit? 
time_stamps %>%  select( sec_in_video , time_in_video, datetime  )



```
# distance between points

### first convert the XY coordinates to utm
Coordinates must be in meters so they are in the same units as depth 

```{r}
# print which utm zone should be used by locating the latitude and longitude of the transects - you can use this website to find out the right UTM zone: https://www.latlong.net/lat-long-utm.html
recomended_utm <- (floor((smoothed_navigation_laser_data$LON2[1] + 180) / 6) %% 60) + 1
print(paste0("the recommended UTM zone for your transects is: ", recomended_utm, " - check if this is correct and change the code if needed!"))

# TO set the utm zone manually: 
  recomended_utm <- 33 # change this to the correct UTM zone for your transects, check with the website mentioned above

# Transform latitude and longitude to UTM coordinates
traj.sf <-st_as_sf(smoothed_navigation_laser_data ,coords =c("LON2","LAT2"),crs="+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")
# re project to UTM 33
# find out the right UTM zone for you transects and convert to it
traj.sf %<>% st_transform( paste0( "+proj=utm +zone=",recomended_utm, " +datum=WGS84 +units=m +no_defs") )
# get coordinates 
traj.sf$geometry %>% st_coordinates() %>% as_tibble() %>% bind_cols(smoothed_navigation_laser_data) -> joint_navigation_annotation_utm

# re-naming smoothed depth vector (DEPTH2) to just depth for simplicity
joint_navigation_annotation_utm %>%
  dplyr::select(X, Y, depth = DEPTH2)  -> dpoint

```


### then calculate the distance

using the dist function from the base R package. It calculate the euclidean distance between two points in 3D space (X, Y, and depth). 

```{r}

# at each point, use coordinates of the previous point 
dpoint %>%
  slice(-n()) %>%  bind_rows(dpoint %>% slice(1),.) %>% 
  rename(X2 = X,Y2 = Y, depth2 = depth) %>%  # USE interpolated depth HERE !!!!!!
  bind_cols(dpoint,.) %>%
  # calculate distance between each point and the point before (so it starts at 0)
  rowwise() %>%
  mutate(distance_3d =
           # put coordinates into a matrix
           matrix(c(X2,Y2,depth2,X,Y,depth), # USE interpolated depth HERE !!!!!!
                  nrow= 2,
                  ncol = 3, # coordinates matrix has  columns
                  byrow = TRUE ) %>% 
           # calculate Euclidean distance between two sets of points 
           dist() %>%  as.numeric()) %>% 
  ungroup() -> dpoint2



#cumulated distance travelled ### remove replace_na 0 later, keep only cumsum(distance_3d) replace 3D with 2D 
dpoint2  %>% 
  mutate(index = 1:nrow(.)) %>% 
  #group_by(index) %>% 
  mutate(distance_travelled = cumsum(distance_3d)) %>% 
  select(distance_travelled ) %>% 
  bind_cols(dpoint2,. ) -> cumulated_3D_distance_travelled


```

# print results and export

```{r}

# print the whole transects length/distance travelled
cumulated_3D_distance_travelled %>% tail(1) %>% pull(distance_travelled) -> v
  paste0("Distance travelled in 3D: ",round(v,2)," meters") %>% 
  print() 
  rm(v)

# store total distance travel as own value for calculating total species density along whole distance in 5.2
cumulated_3D_distance_travelled %>% tail(1) %>% pull(distance_travelled) -> total_distance_travelled
# add the distance column
joint_navigation_annotation <- bind_cols(joint_navigation_annotation_utm,
                                         cumulated_3D_distance_travelled %>% select(distance_3d,distance_travelled))

# export the smoothed and calibrated navigation into the 
joint_navigation_annotation %>%
  write_csv(paste0("./Output/dist_calibrated_smoothed_3d_navigation.csv" ) )

```

```

