---
title: "Analyzing BIIGLE Video Data"
author: "Your Name"
date: "Today's Date"
format:
  html:
    toc: true
    number-sections: true
---

## Introduction

This Quarto document outlines the workflow for analyzing video annotation data from BIIGLE. We will focus on calculating species density changes along a video transect, generating plots, and exporting processed data for further analysis.

## Setup

Before beginning, we will load necessary libraries.

```{r}

library(patchwork)
library(readr)
library(plotly)
library(RColorBrewer)
library(tidyverse)

getwd() -> wd
# delete the "code" part of the pathway so that the rest of the path are not relative to the code directory
wd <- str_remove(wd, "/code") 


# define bins parameters 

# set a bin size in m 
bin_size <- 50



```

## Load navigation data

Next, we load the image calibration data to assist in calculating the seabed surface width.

```{r}
# open image calibration info to calculate the bin surface
paste0(wd,"/Output/dist_calibrated_smoothed_3d_navigation.csv" ) %>%  read_csv() -> navigation_data

```

## Data Binning

To analyze how species density changes along the video transect, we split the dataset into subsets based on the distance travelled by the ROV. Here, we will create 50m distance traveled bins.

```{r}
# create bins of 50m distance travelled 
breaks <- seq(0, max(navigation_data$distance_travelled) + bin_size, by = bin_size)
binned_numbers <- cut(navigation_data$distance_travelled, breaks = breaks, include.lowest = TRUE)
navigation_data <- mutate(navigation_data, bin = binned_numbers) 


navigation_data %>%
  write_csv(paste0(wd,"/Output/bins_dist_calibrated_smoothed_3D_steep3.csv" ) )





```

### calculate bins metadata

```{r}


# show the average width in each bin 
navigation_data %>%  group_by(bin) %>% 
  summarise(n = n(), 
            # caluculate the mean depth in each bin take DEPTH2 for smoothed and only original depth for non-smoothed
            mean_depth = mean(DEPTH2, na.rm = TRUE),
            # and the width
            mean_width = mean(width, na.rm = TRUE),
  ) %>% 
  # surface of a bin is the average width * 50m
  mutate(bin_surface = mean_width * bin_size) -> bins_metadata

# add a line to the tabl showing the average values for the entire table
bins_metadata %>% 
  summarise(mean_depth = mean(mean_depth),
            mean_width = mean(mean_width),
            bin_surface = sum(bin_surface)) %>% 
  mutate(bin = paste0( "All(",breaks %>% head(1),",",breaks %>% last() ,"]"), n = nrow(navigation_data) ) %>% 
  bind_rows(bins_metadata) -> bins_metadata



```

### visualise the bins metadata

```{r}

# make a custom colour vector
nb.cols <- length(breaks)
mycolors <- colorRampPalette(brewer.pal(12, "Set3"))(nb.cols)

bins_metadata %>% 
  # remove the first line (all bins)
  filter(bin != "All(0,950]") %>%
  # make the Bin column a factor
  mutate(bin = factor(bin, levels = unique(bin))) %>%
  ggplot(aes(x = bin, y = - mean_depth, fill = bin)) +
  geom_bar(stat = "identity") +
  # set color palette to "set3"
  scale_fill_manual(values = mycolors) +
   # display x axis labels vertically
   theme_minimal() + 
    # disable the legend
  theme(legend.position = "none") +
  theme (axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Mean Depth by Bin", x = "Bin", y = "Mean Depth (m)") -> plot_bins_metadata

# same plot for the  width
bins_metadata %>% 
  # remove the first line (all bins)
  filter(bin != "All(0,950]") %>%
  # make the Bin column a factor
  mutate(bin = factor(bin, levels = unique(bin))) %>%
  ggplot(aes(x = bin, y = mean_width, fill = bin)) +
  geom_bar(stat = "identity") +
  # set color palette to "set3"
  scale_fill_manual(values = mycolors) +
   # display x axis labels vertically
   theme_minimal() + 
    # disable the legend
  theme(legend.position = "none") +
  theme (axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Mean Width by Bin", x = "Bin", y = "Mean Width (m)") -> plot_bins_metadata_width


# both plots in a grid of 1 row


plot_bins_metadata + plot_bins_metadata_width + 
  plot_layout(ncol = 1) +
  plot_annotation(title = "Mean Depth and Width by Bin") -> plot_bins_metadata_combined

plot_bins_metadata_combined

 

```

## Merge navigation with observation

### Annotation data Loading

Load Data set We will start by loading the output from a previous script that recorded the distance traveled in BIIGLE annotations.

```{r}
# open the table of annotations
arranged_annotations_file <-  paste0(wd,"/annotations/","arranged_annotations_steep3.csv" )
read_csv(arranged_annotations_file) -> arranged_biigle_annotations 
```

We attach the width of the seabed to the dataset using a left join.

```{r}

# first and last frame of each annotation 
arranged_biigle_annotations %<>%
  # start frame
  mutate(startframe =    frames %>% str_sub( 2, -2)  %>% 
           str_replace("null",replacement = "0") %>% 
           str_split(pattern = ",") %>%
           map_vec(~head(.x,1) %>% as.numeric),
         # endframe
         endframe =    frames %>% str_sub( 2, -2)  %>% 
           str_replace("null",replacement = "0") %>% 
           str_split(pattern = ",") %>%
           map_vec(~tail(.x,1) %>% as.numeric) )

# make a start and end second for each annotation
arranged_biigle_annotations %<>%
  mutate( startsec = startframe %>% seconds() %>% floor, 
          endsec = endframe %>% seconds() %>% floor)

arranged_biigle_annotations %>% select(endsec, startsec, frames)  



```

### Join the arranged BIIGLE annotations to the navigation

```{r}


time_stamps <- navigation_data  %>%  
  select(datetime) %>% 
  mutate(sec_in_video = seconds(datetime  -  video_starttime   )  %>% seconds_to_period()  ) %>% 
  mutate(time_in_video =sprintf('%02d:%02d:%02d', sec_in_video$hour,sec_in_video$minute, sec_in_video$second) )
# change format of time column in navigation table
time_stamps %<>% 
  mutate( sec_in_video =  sec_in_video %>% seconds() )
# join to nearest timestamp in metadata  
by <- join_by(  closest(endsec  <= sec_in_video )) # !! joining by closest time stamp - MAY NOT BE EXACT !!!!!!!!!!!
time_stamps  %>%  left_join(arranged_biigle_annotations, . ,by , suffix = c(".annotations", ".metadata"))   -> arranged_biigle_annotations_timestamps


# add the navigation 
# attach the width of the seabed to the dataset exchange DEPTH2 with original depth for non-smoothed
arranged_biigle_annotations_timestamps %>%
  left_join(navigation_data %>% select(X,Y,depth = DEPTH2,width,   datetime , bin ),
            by = c("datetime" = "datetime")) -> dataset

```



## Species Occurrence by Bin

Calculate the number of occurrences of each species within each bin.

```{r}
bins_list <- list()

for (i in unique(dataset$bin)){
  bin.i <- dataset %>% filter(bin == i)
  bin.i.counts <- bin.i %>% count(label_name)
  bin.i.counts$bin <- i 
  bins_list[[i]] <- bin.i.counts
}

bins_df <- bins_list %>% bind_rows()
```

## Pivot Table

Transform the table to have species names as columns and bins as rows.

```{r}
bins_df <- bins_df %>%
  pivot_wider(names_from = label_name, values_from = n) %>% 
  replace(is.na(.), 0)
```

### make a "All" bins first line

```{r}
# add a line to the table showing the average values for the entire table 
bins_df %>% select(-bin) %>% 
  summarise(across(everything(), sum))  %>% 
  mutate(bin =bins_metadata$bin[1] ) %>% 
  relocate(bin, .before = everything()) %>%
  bind_rows(bins_df) -> bins_df

```

## Visualizing

3D Depth Plot We create a 3D plot to visualize the X, Y, and depth dimensions, color-coded by bins.

```{r}
plot_ly(dataset, x = ~X, y = ~Y, z = ~-depth,
        color = ~bin, 
        colors = mycolors,
        type = "scatter3d", mode = "markers", 
        marker = list(size = 2))  %>%
  layout(title = "XY Depth in 3D with Bin Color") 


```

### Exporting Data

Species Abundance Export the table of species abundance per bin.

```{r}
bins_df %>% 
  write_csv(paste0("../Output/",bin_size,"bins_species_abundances_annotations_steep3_3Dsmoothed.csv"))
```

###Species Density Calculate the density of each species in each bin based on surface area.

```{r}
bins_df_density <- bins_df %>%
  select(-bin) %>%
  map(~ .x / bins_metadata$bin_surface) %>%
  bind_cols(bins_df %>% select(bin))  
 
  
# export tables of species densities per bin
bins_df_density %>% 
  write_csv(paste0(wd,"/Output/",bin_size,"bins_species_densities_annotations_steep3_3Dsmoothed.csv"))

### calculate total species density of total transect area
### first calculate total surface from previous calculated average_image_width multiplied with total_distance_travelled value. IMPORTANT: to get the average image width variable, this needs to be first computed in script 3!

total_surface <- total_distance_travelled*average_image_width
print(total_surface)

### select row containing cumulated species occurences along whole transect
total_df_density <- bins_df %>%
  filter(row_number()==1) 

### add column of total surface 
total_df_density %>% mutate(total_surface) -> total_df_density

### calculate total density of each label based on total surface area of whole transect
total_df_density %>% select(-bin) %>%
  map(~ .x / total_surface) %>%
  bind_cols(total_df_density %>% select(bin)) -> total_df_density

### export tables of species density
total_df_density  %>% 
  write_csv(paste0(wd,"/Output/","species_densities_annotations_steep3_3Dsmoothed.csv")) 

```

## Principal Component Analysis (PCA)

Perform PCA to visualize the different bins in the species space.

```{r}

# basinc PCA on the species density data
bins_pca <- bins_df_density %>% 
  # remove the all 
  # filter out hte bin name that has the pattern "All" in it
  filter(!str_detect(bin, "All")) %>%
  # ignore
  select(-bin) %>% 
  prcomp(center = TRUE, scale. = TRUE)

# view results of bins coloured by depth
plot_ly() %>%
  add_trace(x = bins_pca$x[,1], y = bins_pca$x[,2], z = bins_pca$x[,3],
            color = bins_metadata %>% filter(!str_detect(bin, "All")) %>% pull(mean_depth)  ,
            type = "scatter3d", mode = "markers") %>% # 
  layout(title = "PCA of Species Density in 50m Bins",
         scene = list(xaxis = list(title = "PC1"),
                      yaxis = list(title = "PC2"),
                      zaxis = list(title = "PC3")))
# put the name of the bins as labels 

```
